### 2017-10-15 mcmc.tex
g functions

\section{Convergence}\label{sec:convergence}
\nopagebreak
A key question for an \MCMC\ operator---\emph{the} key question in some
sense---is \emph{how long to run} to be sure of having reliable
results.
It is disappointing and annoying to many that there is no extremely
simple and reliable answer to this question.\note{There is very good
  coverage of most of the points in this \sectionname, and more, in a
  set of lecture notes by Patrick Lam up at
  \url{http://www.people.fas.harvard.edu/~plam/teaching/methods/convergence/convergence_print.pdf}.}

@@

In this context,
  it is \emph{simply not fair} to require an \MCMC\ run to have fully and completely
  sampled the posterior pdf,
  at least not in any provable sense.
The question of whether a sampling is converged to a representative sampling
  of the posterior pdf is actually outside the domain of science,
  because the domain of science is the domain of questions answerable by scientists.
That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
  despite many statements in the literature to the contrary.\note{%
For example, you can (almost) never say that your chains are
definitely converged, or that you have the posterior pdf correct to
some given level of accuracy.  The reason is related to the ``no free
lunch'' theorem of discrete optimization: In most problems there is no
way you could have searched your parameter space finely enough to know
this.  There are some exceptions of course.  In one kind of exception,
your problem is convex or Gaussian, and you can analytically show that
there is exactly one mode in the density and where it is.  Of course
in these cases you rarely need \MCMC\ to solve your problems!  In
another kind of exception, you can say something about the finite
width or shape of the modes of the likelihood function (and prior
pdf).  For example, sometimes the Cram\'er--Rao bound tells you that
modes of the density must be smoother than some finite amount.  Then
an exhaustive search of parameter space followed by \MCMC\ can in
principle have provable properties.  But, as we say, these situations
are rare.}
The upshot of this is that we don't and can't require any kind of absolute convergence;
  indeed, it would be impossible even to test for it.

In this pragmatic situation---the Real World, as it is called---we have to rely on heuristics.
Heuristically, you have sampled long enough when you can see
that the (or each) walker has traversed the high-probability parts of
the parameter space many times in the length of the chain.
Or, equivalently, you have sampled long enough when the first half of
the chain shows very much the same posterior pdf morphology as the
second half of the chain, or indeed as any substantial subset.

@@

The autocorrelation function is defined by
\begin{eqnarray}
C_f(\Delta) &\equiv& \frac{E[(f(\pars_k)-\mu_f)\,(f(\pars_{k+\Delta})-\mu_f)]}
                          {E[(f(\pars)-\mu_f)^2]}
\label{eq:autocorrelationfn}\\
\mu_f &=& E[f(\pars)]
\quad ;
\end{eqnarray}
where $f(\pars)$ is a scalar valued function of the parameters.
In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
of points in the chain separated by $\Delta$ steps, taken relative to the
variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C_f(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C_f(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau_f = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_f(\Delta)
\label{eq:autocorrelationtime}\quad.
\end{eqnarray}
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

@@

\item \emph{Corner plots or scatterplot matrices}~---~In a $D$-dimensional
parameter space, we recommend plotting all $D$ one-dimensional sample
histograms and all $D$-choose-2 two-dimensional histograms (scatter plots) to
show the low-level covariances and non-linearities.
If you are clever, all these plots can be arranged into a lovely and
informative triangle.\note{Our favorite package these days is our own
\project{corner.py}.  Find it on the internets.} These two-dimensional plots
do not in any sense contain all the information in the sampling but they are
remarkable for locating expected and unexpected parameter relationships, and
often invaluable for suggesting re-parameterizations and transformations that
simplify your problem.

\end{itemize}

DFM: Add an example of each of these plots.

>>>>

\section{Convergence}\label{sec:convergence}%
\nopagebreak%
A key question for an \MCMC\ operator---\emph{the} key question in some
sense---is \emph{how long to run} to be sure of having reliable
results.
It is disappointing and annoying to many that there is no extremely
simple and reliable answer to this question.\note{There is very good
  coverage of most of the points in this \sectionname, and more, in a
  set of lecture notes by Patrick Lam up at
  \url{http://www.people.fas.harvard.edu/~plam/teaching/methods/convergence/convergence_print.pdf}.}

@@

In this context,
  it is \emph{simply not fair} to require an \MCMC\ run to have fully and completely
  sampled the posterior pdf,
  at least not in any provable sense.
The question of whether a sampling is converged to a representative sampling
  of the posterior pdf is actually \emph{outside the domain of science},
  because the domain of science is the domain of questions answerable by scientists.
That is, you can \emph{never know} that you have correctly sampled the posterior pdf,
  despite many statements in the literature to the contrary.\note{%
For example, you can (almost) never say that your chains are
definitely converged, or that you have the posterior pdf correct to
some given level of accuracy.  The reason is related to the above-mentioned ``no free
lunch'' theorem of discrete optimization \citep{nofreelunch}: In most problems there is no
way you could have searched your parameter space finely enough to know
this.  There are some exceptions of course.  In one kind of exception,
your problem is convex or Gaussian, and you can analytically show that
there is exactly one mode in the density and where it is.  Of course
in these cases you rarely need \MCMC\ to solve your problems!  In
another kind of exception, you can say something about the finite
width or shape of the modes of the likelihood function (and prior
pdf).  For example, sometimes the Cram\'er--Rao bound tells you that
modes of the density must be smoother than some finite amount.  Then
an exhaustive search of parameter space followed by \MCMC\ can in
principle have provable properties.  But, as we say, these situations
are rare.}
The upshot of this is that we don't and can't require any kind of absolute convergence;
  indeed, it would be impossible even to test for it.

In this pragmatic situation---the Real World, as it is called---we have to rely on heuristics.
Heuristically, you have sampled long enough when you can see
that the (or each) walker has traversed the high-probability parts of
the parameter space many times in the length of the chain.
Or, equivalently, you have sampled long enough when the first half of
the chain shows very much the same posterior pdf morphology as the
second half of the chain, or indeed as any substantial subset.
Or, relatedly, you have sampled long enough when different walkers initialized
differently and run with different random number seeds create posterior inferences
that are substantially the same.

@@

The autocorrelation function $C_g$ is defined by
\begin{eqnarray}
C_g(\Delta) &\equiv& \frac{E[(g(\pars_k)-\mu_g)\,(g(\pars_{k+\Delta})-\mu_g)]}
                          {E[(g(\pars)-\mu_g)^2]}
\label{eq:autocorrelationfn}\\
\mu_g &=& E[g(\pars)]
\quad ;
\end{eqnarray}
where $g(\pars)$ is a scalar valued function of the parameters, and $\Delta$
is a lag (difference in index number) between points in the chain.
In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
of points in the chain separated by $\Delta$ steps, taken relative to the
variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C_g(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C_g(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau_g = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_g(\Delta)
\label{eq:autocorrelationtime}\quad.
\end{eqnarray}
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

@@

\item \emph{Corner plots or scatterplot matrices}~---~In a $D$-dimensional
parameter space, we recommend plotting all $D$ one-dimensional sample
histograms and all $D$-choose-2 two-dimensional histograms (scatter plots) to
show the low-level covariances and non-linearities.
If you are clever, all these plots can be arranged into a lovely and
informative triangle.\note{Our favorite package these days is our own
\project{corner.py}.  Find it on the internets.} These two-dimensional plots
do not in any sense contain all the information in the sampling but they are
remarkable for locating expected and unexpected parameter relationships, and
often invaluable for suggesting re-parameterizations and transformations that
simplify your problem.
\end{itemize}
